---
title: "Confidence Region for Bivariate Exponential Mean Vector estimated by EM"
output:
  html_notebook: default
---

```{r}
library(MASS)
library(car)
```

## EM estimation for Bivariate Exponential Distribution

### Creat Bivaraite Exponential Dataset
with theta1 = 2, theta2 = 3, common = 0.6
```{r}
set.seed(12345)
 n <- 1000 # sample size
  lambda1 <- 2
  lambda2 <- 3
  common <- 0.6
  x1 <- rexp(n, rate=lambda1-common)
  x2 <- rexp(n, rate=lambda2-common)
  z <- rexp(n, rate=common)
  y1 <- pmin(x1, z)
  y2 <- pmin(x2, z)
data <- cbind(y1,y2)
```

### E-step: 
Args:
  data.matrix: Matrix containing the bivaraite data
  theta.vector: Vector containing theta1, theta2 and theta0
Return:
  Named list containing z, d, and phi
```{r}
e_step <- function(data.matrix, theta.vector) {
  n <- nrow(data.matrix)
  z <- vector(, n)
  d <- vector(, n)
  phi <- vector(, n)
  theta0 <- theta.vector[3]
  theta1 <- theta.vector[1]; theta2 <- theta.vector[2]
 
   for (row in 1:n){
    y1 <- data.matrix[row,1]; y2 <- data.matrix[row,2]
    if (y1 < y2) {
      z[row] <- y2 + theta2/(theta2+theta0)/theta0
      d[row] <- y1
      phi[row] <- y2 + theta0/(theta2+theta0)/theta2
    }
    if (y1 > y2) {
      z[row] <- y1 + theta1/(theta1+theta0)/theta0
      d[row] <- y1 + theta0/(theta1+theta0)/theta1
      phi[row] <- y2
    }
    if (y1 == y2) {
      z[row] <- y1
      d[row] <- y1 + 1/theta1
      phi[row] <- y2+1/theta2
    }
   
   }
  list('z.vector'=z, 'd.vector'=d, 'phi.vector'=phi)
}
```

### M-step:
Args:
  z.vector
  d.vector
  phi.vector
Return:
 theta.vector: Vector containing theta1, theta2, and theta0
```{r}
m_step <- function(z.vector,d.vector,phi.vector) {
  n <- length(z.vector)
  theta0 <- n/sum(z.vector)
  theta1 <- n/sum(d.vector)
  theta2 <- n/sum(phi.vector)
  theta.vector <- c(theta1,theta2,theta0)
  return (theta.vector)
}
```

###Main function of EM algorithm:
Args:
  data
  initial.theta.vector
Return:
  theta.vector containing converged theta1, theta2 and theta0

```{r}
em_main <- function(data, initial.theta) {
  for (i in 1:1000) {
    if (i ==1) {
    # Initialization
    e.step <- e_step(data, initial.theta)
    theta <- m_step(e.step$z.vector,e.step$d.vector,e.step$phi.vector)
    }
    
    else {
    # Repeat E and M steps till convergence
      theta_prev <- theta
      e.step <- e_step(data, theta)
      theta <- m_step(e.step$z.vector,e.step$d.vector,e.step$phi.vector)
      #Convergence check
      if (identical(theta_prev, theta)){
        break
      }
    }
  }
  lamda <- c(theta[1]+theta[3],theta[2]+theta[3],theta[3])
  return(lamda)
}
```

```{r cache=TRUE}
theta.initial <- c(1, 1, 0.5)
em_parameter <- em_main(data, theta.initial)
em_parameter
cat ("Estimated lambda1 is:", em_parameter[1],"\n")
cat ("Estimated lambda2 is:", em_parameter[2],"\n")
cat ("Estimated common is:", em_parameter[3])
```


## Confidence Region for lambda1 and lambda2 in Bivariate Exponential Distribution
###T2 Method



###Parametric Bootstramp

#### Step1: Function to Generate Bivaraite Exponential 
```{r}
BiExp <- function (lambda1, lambda2, common) {
  n <- 1000 # sample size
  x1 <- rexp(n, rate=lambda1-common)
  x2 <- rexp(n, rate=lambda2-common)
  z <- rexp(n, rate=common)
  y1 <- pmin(x1, z)
  y2 <- pmin(x2, z)
  return (cbind(y1,y2))
}
```

#### Step2: Parameter Bootstramp for Bi-Exponential
```{r}
exp.bs <- function(x,y){
  #Step0: find em estimators 
  data <- cbind(x,y)
  theta.initial <- c(1, 1, 0.5)
  lambda.vector <- em_main(data, theta.initial)
  #sample with em estimators one time
  BiExp.lambda <- function (lambda1, lambda2, common){
    #Step1: sample from Bi-Exponential with estimated parameter
    sample <- BiExp(lambda1, lambda2, common)
    
    #Step2: Compute the estimated lambdas in this sample
    initial <- c(1, 1, 0.5) #initial parameter for em estimation in each bootstrap sample
    lambda.sample <- em_main(sample, initial)
    return (lambda.sample)
  }
  
  #repeat sampling 200 times
  simlist <- replicate(200, BiExp.lambda(lambda.vector[1],lambda.vector[2], lambda.vector[3]))
  return (t(simlist))
}

x <- data[,1]
y <- data[,2]
para_bootstrap = exp.bs(x,y) 

```

#Parametric Bootstrap Results
```{r}
print (para_bootstrap)
```

##Non-parametric Bootstramp
```{r}
#library(MASS)
nonp.bs <- function(x,y){
  
  sample.lambda <- function(x,y){
    #Step1: sample with raplacement directly from data
    n = length(x)
    data = cbind(x,y)
    index = sample(1:n, size=n, replace=TRUE)
    sample <- data[index,]
    #Step2: Compute estimated lambdas in this sample
    initial <- c(1, 1, 0.5) #initial parameter for em estimation in each bootstrap sample
    lambda.sample <- em_main(sample, initial)
    return (lambda.sample)
  }
  simlist <- replicate(200, sample.lambda(x,y))
  return (t(simlist))
}
nonp_bootstrap = nonp.bs(x,y)
```

```{r}
print (nonp_bootstrap)
```

## Hoetelling T-squared method
###This is wrong because the stats does't follow F distribution in this case. The results are just for comparision
```{r}
t2.r<- function(x,y){
  p <- 2
  n <- length(x)
  radius <- p*(n-p)/(n-1)*qf(0.95, p, n-p)
  return (sqrt(radius))
}
t2.radius <- t2.r(x,y)

```


## Ploting Confidence Region
### Green line is parametric, while blue line is nonparametric
```{r}
library(car)

#Compute the covariance matrix for T2 with parametric_bootstrap results
s = cov(para_bootstrap)
s.partial = s[1:2,1:2]

plot(x,y,pch='.',xlim=c(1.5, 2.5), ylim=c(2.5, 3.5),main="Confidence Region for lambdas(2,3) in Bivaraite Exponential" )
ellipse (center = em_parameter[1:2], shape = s.partial, radius = t2.radius, draw=TRUE, col = 'red') 
dataEllipse(para_bootstrap[,1], para_bootstrap[,2], levels = c(0.95), col='blue', add=TRUE, pch='.')
dataEllipse(nonp_bootstrap[,1], nonp_bootstrap[,2], levels = c(0.95), col='green', add=TRUE, pch='.')
legend("topleft", legend=c("T squared", "Parametric Bootstrap", "Nonparametric Bootstrap"),
       col=c("red","blue","green"),lty=1)
```

```{r}

```

