---
title: "Comparing EM and K-means with GMM"
output: html_notebook
---


```{r}
library(dplyr)  #to organize kmeans result, used group_by(), summarize()
library(ggplot2) #to plot results
library(ggplot2) #to plot histgram and prob density
library(mixtools) #EM algorithm 
```



##E-step：Expectation Step of the EM Algorithm
Calculate the posterior probabilities (soft labels) that each component has to each data point.
Args:
  sd.vector: Vector containing the standard deviations of each component
  sd.vector: Vector containing the mean of each component
  weight.vector: Vector containing the mixing weights  of each component
Return:
  Named list containing the loglik and posterior.df
```{r}
e_step <- function(x, mu.vector, sd.vector, weight.vector) {
  comp1.prod <- dnorm(x, mu.vector[1], sd.vector[1]) * weight.vector[1]
  comp2.prod <- dnorm(x, mu.vector[2], sd.vector[2]) * weight.vector[2]
  comp3.prod <- dnorm(x, mu.vector[3], sd.vector[3]) * weight.vector[3]
  sum.of.comps <- comp1.prod + comp2.prod + comp3.prod
  comp1.rho <- comp1.prod / sum.of.comps
  comp2.rho <- comp2.prod / sum.of.comps
  comp3.rho <- comp3.prod / sum.of.comps
  sum.of.comps.ln <- log(sum.of.comps, base = exp(1))
  
  list("loglik" = sum(sum.of.comps.ln),
       "rho.df" = cbind(comp1.rho, comp2.rho, comp3.rho))
}
```


##M-Step: Update the Component Parameters
Arg:
  x Input data.
  rho.df Posterior probability data.frame.
Return:
  Named list containing the mean (mu), variance (var), and mixing weights (weight) for each component.
```{r}
m_step <- function(x, rho.df) {
  comp1.n <- sum(rho.df[, 1])
  comp2.n <- sum(rho.df[, 2])
  comp3.n <- sum(rho.df[, 3])
  
  comp1.weight <- comp1.n / length(x)
  comp2.weight <- comp2.n / length(x)
  comp3.weight <- comp3.n / length(x)
  
  comp1.mu <- sum(rho.df[, 1] * x)/comp1.n
  comp2.mu <- sum(rho.df[, 2] * x)/comp2.n
  comp3.mu <- sum(rho.df[, 3] * x)/comp3.n
  
  comp1.var <- sum(rho.df[, 1] * (x - comp1.mu)^2)/comp1.n
  comp2.var <- sum(rho.df[, 2] * (x - comp2.mu)^2)/comp2.n
  comp3.var <- sum(rho.df[, 3] * (x - comp3.mu)^2)/comp3.n

  list("mu" = c(comp1.mu, comp2.mu, comp3.mu),
       "var" = c(comp1.var, comp2.var, comp3.var),
       "weight" = c(comp1.weight, comp2.weight, comp3.weight))
}
```


##Main function of EM algorithm:
Iterate through e-step and m-step, 1000 iterations in total
Stop if the loglikelihood increment is less than 1e-6

Args:
  x,
  initial.df, initial mu, sigma and weight
Return:
  Named list containing the converged mean (mu), variance (var), and mixing weights (weight) for each component,
  as well as # of iterations, and log-likelyhood vector.
```{r}
em_main <- function(x, initial.df){
  for (i in 1:1000) {
    print (i)
    if (i == 1) {
    # Initialization
      e.step <- e_step(x, initial.df[["mu"]], initial.df[["std"]], initial.df[["weight"]])
      m.step <- m_step(x, e.step[["rho.df"]])
      prev.loglik <- e.step[["loglik"]]
      loglik.vector <- e.step[["loglik"]]
      }
    else {
    # Repeat E and M steps till convergence
      #E-step
      e.step <- e_step(x, m.step[["mu"]], sqrt(m.step[["var"]]), m.step[["weight"]])
      #M-step
      m.step <- m_step(x, e.step[["rho.df"]])
      #Convergence check
      loglik.vector <- c(loglik.vector, e.step[["loglik"]])
      loglik.diff <- e.step[["loglik"]]-prev.loglik
      if(loglik.diff < 1e-6) {
        break
        } 
      else {
        prev.loglik <- e.step[["loglik"]]
        }
      }
    }
  list("mu" = m.step[["mu"]],
       "var" = m.step[["var"]],
       "weight" = m.step[["weight"]],
       "number of iterations" = length(loglik.vector),
       "loglik" = loglik.vector)
}
```


Part 0: Generate Data
Generate a sample of size n = 150 from normal mixture: 
p1*N(-5; 4) + p2*N(0; 0.25) + p3*N(2, 1) with p1 = 0.5, p2 = 0.3, and p3 = 0.2 

```{r}
set.seed(1)
N <- 150
components <- sample(1:3,prob=c(0.5,0.3,0.2),size=N,replace=TRUE)
mus <- c(-5, 0, 2)
sds <- sqrt(c(4, 0.25, 1))
samples <- rnorm(n=N,mean=mus[components],sd=sds[components])
samples
```


Part 2. K-means clustering 
```{r}
samples.kmeans <- kmeans (samples, 3)
samples.kmeans
```

```{r}
samples.kmeans.cluster <- samples.kmeans$cluster
samples.df <- data.frame (x = samples, lable = components, kmeans = samples.kmeans.cluster)
samples.df
```
Compute weight
```{r}
samples.summary.df <- samples.df %>%
  group_by(kmeans)  %>%
  summarize(mu = mean(x), std = sd(x), size = n())

kmeans.summary.df <- samples.summary.df  %>%
  mutate(weight = size / sum(size))
kmeans.summary.df
```



Part 1.1, EM algorithm
Use KMeans result as initialization

```{r}
em.result <- em_main(x = samples, initial.df = kmeans.summary.df)
em.result
```


part1.3  EM in mixtools 
```{r}
mixmdl <- normalmixEM(samples, k = 3)
print (mixmdl$mu)
print (mixmdl$sigma)
print (mixmdl$lambda)
mix.var <-  mixmdl$sigma^2
print (mixmdl)

```

Part 1.2 plotting
```{r}

true.weight = c(0.5, 0.3, 0.2)
true.mu = c(-5, 0, 2)
true.var = c(4, 0.25, 1)

plot_mix_comps <- function(x, mu, var, weight){
  density <- 0
  for (i in 1:3){
    density <- density + weight[i]*dnorm(x,mu[i],sqrt(var[i]))
  }
  return(density)
}

ggplot(data.frame(samples), aes(samples)) +
  geom_histogram(aes(samples, ..density..), binwidth = 0.1, colour = "black", fill = "white" ) +
  #geom_density() + 
  stat_function(geom = "line", fun = plot_mix_comps, colour = "red", lwd = 1,
                               args = list(true.mu, true.var, true.weight)) +
  stat_function(geom = "line", fun = plot_mix_comps, colour = "green", lwd = 1,
                               args = list(mixmdl$mu, mix.var, mixmdl$lambda)) +
  stat_function(geom = "line", fun = plot_mix_comps, colour = "blue", lwd = 1,
                               args = list(em.result$mu, em.result$var, em.result$weight)) +
  geom_text(label = "True distribution", x = 2, y = 0.45, color = "red") +
  geom_text(label = "EM estimate", x = 2, y = 0.4, color = "blue") +
  geom_text(label = "EM by mixtools", x = 2, y = 0.35, color = "green")
  
```




```{r}
library(MASS)
Sigma <- matrix(c(8,3,3,2),2,2)
Sigma
set.seed(12345)
data1 <- mvrnorm(n = 50, rep(-2, 2), Sigma)
data2 <- mvrnorm(n = 50, rep(-2, 2), Sigma)


```

